---
title: Week 1 - Initial Analyses
author: Package Build
date: '2024-09-07'
slug: week-1-initial-analyses
categories: []
tags: []
---

In this first blog post, I begin by conducting some initial analyses of the American elections. Using data from 1948-2020, I analyze how competitive elections are and the proportion of states that vote blue/red. My thanks to Matt Dardet for writing much of this code during and before our weekly section, as well as ShuXin Ho for helping me identify latitude/longitude loading bugs in my maps.



We began by producing a general line plot showing trends in two-party vote share for each party between 1948 and 2020, as well as a map of state election winners by year. Both of these are demonstrated below

```{r}
library(ggplot2)
library(maps)
library(tidyverse)

# Load in data
d_popvote <- read_csv('popvote_1948-2020.csv')

# Create pivoted dataset and add winner column
d_popvote_wide <- d_popvote |>
  select(year, party, pv2p) |>
  pivot_wider(names_from = party, values_from = pv2p) |>
  mutate(winner = ifelse(democrat > republican, "D", "R"))

```

```{r}
line_plot_win_trend <- d_popvote |>
  ggplot(mapping = aes(x = year, y = pv2p, color = party)) +
  geom_line() +
  scale_color_manual(values = c("dodgerblue4", "firebrick1")) +
  theme_bw() +
  labs(title = "Percentage of Popular Vote by Party, 1948-2020",
       x = "Year",
       y = "Two-Party Vote Percentage",
       color = "Party")

states_map <- map_data("state")


```



